Physcraper: Body

### 2020-06-26 2-body.Rmd
comments on everythin except the xamples

  Users can provide a search taxon that is either a more or a less inclusive
  clade relative to the ingroup of the original phylogeny. If the search taxon is more inclusive, the sequence search will be performed outside the MRCAT of the matched taxa, e.g., including all taxa within
  the family or the order that the ingroup belongs to. If the search taxon is a less inclusive clade, the users can focus on enriching a particular clade/region within the ingroup of the phylogeny.
- The Basic Local Alignment Search Tool, BLAST [@altschul1990basic; @altschul1997gapped] is used to identify
  similarity between DNA sequences within the search taxon in a nucleotide
  database, and the sequences on the checked alignment.
  The `blastn` function from the BLAST command line tools [@camacho2009blast] is used for local-database sequence searches.
  For remote-database searches, we modified the BioPython [@cock2009biopython] BLAST function in the [NCBIWWW module](https://biopython.org/DIST/docs/api/Bio.Blast.NCBIWWW-module.html) to accept an alternate BLAST URL. This is useful when a user has no access to the computer capacity needed to setup a local database, and a local blast database can be set up on a remote machine to BLAST without NCBI's required wait times.
  <!--***How does the blast is setup? Is it an all-blast-all??***-->
- A pairwise BLAST search is performed. This means that each sequence
  in the alignment is BLASTed against DNA sequences in a nucleotide database constrained to the search
  taxon. Results from each one of these BLAST runs are recorded, and matched sequences are saved
  along with their corresponding identification numbers (accession numbers in the case of the GenBank database). This information will be used later to store the whole sequences in a dedicated library within the physcraper folder, allowing for secondary analyses to run significantly faster.
- Matched sequences will be discarded if the fall below a default e-value of 0.00001, and outside a default minimum and maximum length of 80% and 120%, respectively, of the average length of sequences in the checked alignment (gaps dropped). 
These parameters can be configured for each run.
  This filtering guarantees that whole genome sequences are not included, as they 
  All accepted sequences are assigned an internal identifier, and are further filtered.
- Because the original alignments usually lack taxonomic identification numbers, an extra filtering
  step is needed. Accepted sequences that belong to the
  same taxon of the query sequence, and that are either identical or shorter than
  the original sequence are discarded. Only longer sequences belonging to the
  same taxon as the orignal sequence will be considered further for analysis.
- Among the remaining filtered sequences, there are usually several exemplars per taxon.
  Although it can be useful to keep some of them, for example, to investigate monophyly
  within species, there can be hundreds of exemplar sequences per taxon for some markers.
  To control the number of sequences per taxon in downstream analyses,
  5 sequences per taxon are chosen at random. This number is set by default but can be modified by the user.
- Reverse complement sequences are identified and translated.
- Users can choose to perform a more "cycles" of sequence similarity search, by blasting the newly found sequences. This can be done iteratively, but by default only sequences in the checked alignment are blasted. I ALSO FORGOT TO ASK: ***Is there an argument to control the number of cycles of blast searches with new sequences?*** \comment(there's only two options - one cycle or, -r where it runs until it doesn't find anything new)
- Accepted sequences are downloaded in full, and stored as a local database in a directory that is globally accessible (default to physcraper/taxonomy), so they are accessible for further runs.
- A fasta file containing all filtered and processed sequences resulting from the BLAST search is generated for the user.

@@

- The software MUSCLE [@edgar2004muscle] is implemented to perform sequence alignments.
- First, all new sequences are aligned using default MUSCLE options.
- Then, a MUSCLE profile alignment is performed, in which the original alignment
  is used as a template to align new sequences. This ensures that the final alignment
  follows the homology criteria established by the original alignment.
- The final alignment is not further processed automatically. So, we encourage users
  to check it either by eye and perform manual refinement or using any of the many
  tools for alignment processing, to eliminate columns with no information.

@@

- A gene tree is reconstructed for each alignment provided, using a Maximum Likelihood approach implemented with the software RAxML [@stamatakis2014raxml], using default settings such as a GTRCAT model of molecular evolution and 100 bootstrap replicates with default method. Currently only the number of bootsrap replicates can be modified by the user.
- The original tree is used as starting tree for the ML searches. It can also be set as a full topological constraint or not used at all, depending on the goals of the user.
- Bootstrap results are summarized with Dendropy [@sukumaran2010dendropy]
- The final result is an updated phylogenetic hypothesis for each of the genes provided in the alignment.
- Tips on all trees generated by physcraper are defined by a taxon name space, allowing to perform comparisons and conflict analyses.
- Robinson Foulds weighted and unweighted metrics are calculated with Dendropy functions.
- Finally a conflict analysis is performed. This is basically a node by node comparison between the the synthetic OpenTree and the original and updated tree individually. This is performed with OpenTree's conflict Application Programming Interface [@redelings2017supertree].
- For the conflict analysis to be meaningful, the root of the tree ineeds to be accurately defined.
- A suggested default rooting based on OpenTree's taxonomy is implemented for now. It takes one descendant from the corresponding induced subtree in OpenTree's synthetic tree and then finds the MRCA of those in th eupdated trees. However, if the updated tree changes expectations from taxonomy, the root will no longer be accurate. Automatic identification of a phylogenetic tree root is indeed a difficult problem that has not been solved yet. The best way right now is for users to define outgroups so trees are accurately rooted.

>>>>

  Users can provide a search taxon that is either a more or a less inclusive
  clade relative to the ingroup of the original phylogeny. If the search taxon is more inclusive, the sequence search will be performed outside the MRCAT of the matched taxa, e.g., including all taxa within
  the family or the order that the ingroup belongs to. If the search taxon is a less inclusive clade, the users can focus on enriching a particular clade/region within the ingroup of the phylogeny.
- The Basic Local Alignment Search Tool, BLAST [@altschul1990basic; @altschul1997gapped] is used to identify
  similarity between DNA sequences within the search taxon in a nucleotide
  database, and the sequences on the checked alignment.
  The `blastn` function from the BLAST command line tools [@camacho2009blast] is used for local-database sequence searches.
  For remote-database searches, we modified the BioPython [@cock2009biopython] BLAST function in the [NCBIWWW module](https://biopython.org/DIST/docs/api/Bio.Blast.NCBIWWW-module.html) to accept an alternate BLAST URL. This is useful when a user has no access to the computer capacity needed to setup a local database, and a local blast database can be set up on a remote machine to BLAST without NCBI's required wait times.
  <!--***How does the blast is setup? Is it an all-blast-all??***-->
- A pairwise BLAST search is performed. This means that each sequence
  in the alignment is BLASTed against DNA sequences in a nucleotide database constrained to the search
  taxon. Results from each one of these BLAST runs are recorded, and matched sequences are saved
  along with their corresponding identification numbers (accession numbers in the case of the GenBank database). This information will be used later to store the whole sequences in a dedicated library within the physcraper folder, allowing for secondary analyses to run significantly faster.
- Matched sequences will be discarded if the fall below a default e-value of 0.00001, and outside a default minimum and maximum length of 80% and 120%, respectively, of the average length of sequences in the checked alignment (gaps dropped). 
These parameters can be configured for each run.
  This filtering guarantees that whole genome sequences are not included, as they 
  All accepted sequences are assigned an internal identifier, and are further filtered.

- New sequences that are identical to existing sequences, or to subsets of an existing sequence are discarded, unless they reperesent a different taxon than the existing sequence.
Longer sequences belonging to the
  same taxon as the original sequence will be considered further for analysis.
- Among the filtered sequences, there are often several representatives per taxon.
  Although it can be useful to keep some of them, for example, to investigate monophyly
  within species, there can be hundreds of exemplar sequences per taxon for some markers.
  To control the number of sequences per taxon in downstream analyses,
  5 sequences per taxon are chosen at random. This number is set by default but can be modified by the user.
- Reverse complement sequences are identified and translated.
- Users can choose to perform a more "cycles" of sequence similarity search, by blasting the newly found sequences. This can be done iteratively, but by default only sequences in the checked alignment are blasted. I ALSO FORGOT TO ASK: ***Is there an argument to control the number of cycles of blast searches with new sequences?*** \comment(there's only two options - one cycle or, -r where it runs until it doesn't find anything new)
- Accepted sequences are downloaded in full, and stored as a local database in a directory that is globally accessible (default to physcraper/taxonomy), so they are accessible for further runs.
- A fasta file containing all filtered and processed sequences resulting from the BLAST search is generated for the user, and is used as an input for alignment.

@@

- The software MUSCLE [@edgar2004muscle] is used by default to perform sequence alignments.
- First, all new sequences are aligned using default MUSCLE options.
- Then, a MUSCLE profile alignment is performed, in which the original alignment
  is used as a template to align new sequences. This ensures that the final alignment
  follows the homology criteria established by the original alignment.
- The final alignment is not further processed automatically. So, we encourage users
  to check it either by eye and perform manual refinement or using any of the many
  tools for alignment processing, to eliminate columns with no information.
- Users may also use physcraper only to gather sequence matches, and apply their own preferred alignemnt and phylogenetic inference tools.

@@

- A gene tree is reconstructed for each alignment provided, using a Maximum Likelihood approach implemented with the software RAxML [@stamatakis2014raxml], using default settings such as a GTRCAT model of molecular evolution and 100 bootstrap replicates with default method. Currently only the number of bootsrap replicates can be modified by the user. 
- The original tree is used as starting tree for the ML searches. It can also be set as a full topological constraint or not used at all, depending on the goals of the user.
- Bootstrap results are summarized with Dendropy [@sukumaran2010dendropy].
- The final result is an updated phylogenetic hypothesis for each of the genes provided in the alignment.
- Tips on all trees generated by physcraper are defined by a taxon name space, capturing the NCBI accession information, as well as the taxon identifiers, allowing the user perform comparisons and conflict analyses.
- Robinson Foulds weighted and unweighted metrics between the tips in the input tree and those tips in the updated tree are calculated with Dendropy functions [@sukumaran2010dendropy].
- Finally a conflict analysis is performed. This is basically a node by node comparison between the the synthetic OpenTree and the original and updated tree individually. This is performed with OpenTree's conflict Application Programming Interface [@redelings2017supertree].
- For the conflict analysis to be meaningful, the root of the tree needs to be accurately defined.
- A suggested default rooting based on OpenTree's taxonomy is implemented for now. This approach uses the taxon labels for all the tips in the updated tree, pulls an inferred subtree from OpenTree's taxonomy and then applies the same rooting to the inferred updated tree. However, if the updated tree changes expectations from taxonomy, the root may no longer be appropriate. Automatic identification of a phylogenetic tree root is indeed a difficult problem that has not been solved yet. The best way right now is for users to define outgroups so trees are accurately rooted.
